A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-20 paper.tex
edited summary

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  considered as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show in the limit of perfect coupling, that  dust-settling cannot lead to axisymmetric instabilities, however thin the dust layer become.     
%
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is similar to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  considered as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show in the limit of perfect coupling, that  dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
% 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We discuss thermodynamical interpretations of 
dust-gas drag instabilities in disks, for example the streaming
instability, as a phase lag between the pressure 
and density of a dusty gas, which is   
analogous to pulsational instabilities 
in stars.  
\end{abstract}

### 2016-08-20 summary.tex
edited summary

%give numerical example 
We have applied our one-fluid stability framework to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by \citeauthor{loren15}, as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity $Z=0.01$ and dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The disk parameters are $p=-1.5,\, q=-1$, and  
$\hgas=0.05$. We set $k = \pi/H_\mathrm{d}$.

Fig. \ref{result2d_loren} shows an unstable mode found for this setup,
with  
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. This is not a VSI mode, as we
find it persists when $q=0$. We estimate a much smaller growth
rate for the streaming instability, of 
$O(10^{-3}\hgas\OmK)$, for the above parameters. We thus also rule out
SI, as did \citeauthor{loren15}.

@@

We obtain the Solberg-Hoiland criteria to assess 
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf dust settling, in itself, cannot lead to
  axisymmetric instability, however large the associated vertical
  shear} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  in dusty disks][]{lee10}. We show that
sharp edges in the dust-to-gas ratio cannot persist in protoplanetary
disks, as these imply a sharp entropy gradient and may thus destablize
the disk.

>>>>

%give numerical example 
We have applied our one-fluid stability analysis to find unstable modes  
similar to that described by \cite{loren15}. We use the same
parameters reported by them as follows. 
%we set $p=q=0$, which eliminates VSI and minimizes streaming
%instabilities as there is no radial pressure gradient at the disk
%midplane.   
The total metalicity is $Z=0.01$ and the dust scale-height $H_\mathrm{d} =
0.4H_\mathrm{g}$, giving a maximum dust-to-gas ratio
$\mathrm{max}(\epsilon) = 0.025$. The midplane stopping time is 
$t_\mathrm{s0}=0.02\OmK^{-1}$. The disk parameters are $p=-1.5,\, q=-1$, and  
$\hgas=0.05$. We set $k = \pi/H_\mathrm{d}$.

Fig. \ref{result2d_loren} shows an unstable mode found for this setup,
with  
frequency 
\begin{align*}
\sigma = (0.3631\ii - 0.1251)\hgas\OmK. 
\end{align*}
This growth rate corresponds to $8.8$ orbits, comparable to the $12$
orbits reported by \citeauthor{loren15}. We find this mode 
persists when $q=0$, so it is not the VSI.  For the above parameters
we also estimate a much smaller growth
rate for the streaming instability (using
Eq. \ref{streaming_dispersion}) of 
$O(10^{-3}\hgas\OmK)$, We thus also rule out
SI, as did \citeauthor{loren15}.

@@

We obtain the Solberg-Hoiland criteria to assess 
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf the vertical shear associated with thin dust
 layers, however large it is, cannot lead to axisymmetric
  instabilities} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  in dusty disks, ][]{lee10}. We show that
sharp edges in the dust-to-gas ratio cannot persist in protoplanetary
disks, as these imply a sharp entropy gradient and may thus destablize
the disk.
